---
title: "R Notebook"
output: html_notebook
date: 14-6-2025
author: Nathan Williams
---

First job is to install and load the packages that we are going to use today. A very useful package that many people use in R is tidyverse. It contains many useful tools such as dyplr for data wrangling, ggplot for making figures etc. IF you'd like to read more about what is available in this package, please see: https://www.tidyverse.org/

```{r Install and load packages}
install.packages("tidyverse") #This is only required once.
library("tidyverse")
```
 
The next job is to set our working directory! If you have saved the GRUMP data from CMAP to your desktop in a folder called "GRUMP-Workshop", then you should be good to go.

```{r Set Working directory}
setwd("~/Desktop/GRUMP-Workshop")
grump_data <- read_csv("Grump_Data.csv")
```

OK! We have the dataset loaded into R studio! Now, the first thing to do is familarise yourself with the dataset. Using str() will enable you to see whats in each column and its name.

```{r Take a quick look at the dataset}
str(grump_data)
```

```{r}

```

